---
title: "Data Analytics I, In Class Exercises"
author: "QBS Bootcamp 2025"
output: pdf_document
---

# Functions and Loops

1) For questions 1 - 3, you will need the data files in the "in_class_exercise_data" zip file we emailed you last night. Save this zip file as a folder on your local computer. Write a function that takes the directory as an input, loops through all the files in the directory, and identifies individuals with CRP levels greater than 0.4mg/dL. Write the data for these individuals with high CRP to fiver separate files, one for each sample data file.
```{r setup, include=TRUE}

```

2) Write a function that takes the directory and generates one set box plots for male and female CRP levels for all five sets of sample data.
```{r}

```

3) Write a function that takes the directory and plots a scatterplot of age and crp for each sample data set. You should end up with five scatter plots. Color the scatter plot dots  by disease status. 
```{r}

```

# Functions and Table One

4) Design your own function to build a table one. Users should input a data frame and how different variables should be treated (i.e. normally or non-normally distributed, categorical).
```{r}

```

